---
title: "RQA sensitivity analysis skin temperature(radius = 0.01)"
author: "Nikolaos Efthymiou"
date: "16/11/2022"
output:
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r initialization, include=FALSE}
rm(list = ls(all = TRUE)) # Clean Enviroment
library(tidyverse)        # A collection of R packages designed for data science
library(magrittr)         # A Forward-Pipe Operator for R
library(knitr)            # A General-Purpose Package for Dynamic Report Generation in R
library(readxl)           # Makes it easy to get data out of Excel and into R
library(nonlinearTseries) # Nonlinear Time Series Analysis
library(glue)             # Interpreted String Literals
library(fs)               # Cross-Platform File System Operations Based on 'libuv'
library(lubridate)        # Time related quirks
library(factoextra)       # Extract and Visualize the Results of Multivariate Data Analyses
library(FactoMineR)       # Multivariate Exploratory Data Analysis and Data Mining
library(lmerTest)         # Tests in Linear Mixed Effects Models
library(tidymodels)       # A collection of packages for modeling and statistical analysis
library(broom.mixed)      # Tidying Methods for Mixed Models
library(ggpubr)           # 'ggplot2' Based Publication Ready Plots
library(formula.tools)    # Programmatic Utilities for Manipulating Formulas, Expressions, Calls, Assignments and Other R Objects
library(lubridate)        # Time related quirks
library(lvmisc)           # Veras Miscellaneous
library(table1)           # An R package for generating tables of descriptive statistics in HTML.
library(gridExtra)        # Miscellaneous Functions for "Grid" Graphics

library(conflicted)  # An Alternative Conflict Resolution Strategy
conflict_prefer("filter", "dplyr")
conflict_prefer("lmer", "lmerTest")
knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, results='asis')
# reproducible results
set.seed(1)
options(ggrepel.max.overlaps = Inf)
```

```{r load-data}
basic <- read_delim("data/Recruitment Management Form-use in R.csv"       , show_col_types = FALSE)
baseline <- read_delim("data/Demographics.csv"                            , show_col_types = FALSE) 
chrono <- read_delim("data/Chronotype_Revised_Theo.csv"                   , show_col_types = FALSE) 
code_setting <- read_delim("data/Recruitment Management Form-use in R.csv", show_col_types = FALSE) 

# load measurements of adiponectin, cortisol, leptin, creatinine and sample types
creatinine     <- read_delim("data/Creatinine results only - final.csv", show_col_types = FALSE)
adiponectin    <- read_delim("data/adiponectin final meas.csv"         , show_col_types = FALSE) 
leptin         <- read_delim("data/leptin final meas.csv"              , show_col_types = FALSE)
cortisol       <- read_delim("data/cortisol final meas.csv"            , show_col_types = FALSE)
SampleTypedata <- read_delim("data/samples.csv"                        , show_col_types = FALSE) 

# read all csv files that contain "S". S means skin data
temperature_measurements <- dir_ls("data/temperature data/", regexp = "S") %>% 
  map(~quietly(read_delim)(.,show_col_types = FALSE, col_types  = "dcddddl")$result)
```

```{r data-modification}
# FSSD to FMS
SampleTypedata$SampleType %<>% {if_else(.=="FSSD","FMS",.)}

# group: Urban-Rural=1 and Rural-Urban=2
code_setting %<>% 
  select(Code,Group) %>%
  bind_rows(.,.) %>%
  group_by(Code) %>%
  transmute(Code_order=glue("{Code}-{1:2}"), setting=if_else(Group==1,c("Urban","Mountainous"),c("Mountainous","Urban"))) %>%
  ungroup() %>%
  select(!Code)

measurements <- 
  creatinine %>%                                         # creatinine measurments 
  mutate(ID=glue("T{str_sub(ID)}"))                  %>% # add char T in front to match with other ID's ex. "65-2-3" -> "T65-2-3"
  full_join(adiponectin   , by=c("ID"="sampleID"))   %>% # adiponectin measurments 
  full_join(leptin        , by=c("ID"="sampleID"))   %>% # leptin measurments
  full_join(cortisol      , by=c("ID"="sampleID"))   %>% # cortisol measurments
  left_join(SampleTypedata, by=c("ID"="sampleID"))   %>% # sample type (First morning sample & Before sleep)
  mutate(ID=str_sub(ID,1,5))                         %>% # drop sample number ex. "T65-2-3" -> "T65-2"
  inner_join(code_setting , by=c("ID"="Code_order")) %>% # add setting (Mountainous-Urban)
  mutate(ID=glue("{str_sub(ID,1,4)}{if_else(setting==\"Urban\",\"U\",\"M\")}")) %>% # include setting in the ID ex. "T10-1" -> "T10-1-U"
  select(!setting) # drop setting 

rm(creatinine,adiponectin,leptin,cortisol,SampleTypedata)
```

```{r temperature-data-filtering-and-loading}
# file name as code
names(temperature_measurements) %<>% str_match("T\\d\\d_S\\d") %>% {glue("{str_sub(.,1,3)}-{str_sub(.,6,6)}")}

# drop T38 because of faulty biomarker measurments 
keep    <- temperature_measurements %>% names() %>% str_sub(1,3) %>% table() %>% as_tibble() %>% filter(n==2 & !(. == "T38"))   %>% select(-n) %>% unlist()
nokeep  <- temperature_measurements %>% names() %>% str_sub(1,3) %>% table() %>% as_tibble() %>% filter((n==1 & !(. == "T38"))) %>% select(-n) %>% unlist()
keepall <- temperature_measurements %>% names() %>% str_sub(1,3) %>% table() %>% as_tibble()                                    %>% select(-n) %>% unlist()

temperature_measurements <- temperature_measurements[temperature_measurements %>% names() %>% str_sub(1,3) %in% keep]
total  <- length(temperature_measurements)

temperature_measurements %<>% 
  bind_rows(.id = "ID")          %>% # bind tibbles
  select(c(ID,Temperature,Date)) %>% # keep only useful columns
  na.omit()                          # drop na

# identify participants with faulty skin temperature measurements (80% of temperature measurements bellow 30c)
faulty_temp_measurements <- temperature_measurements %>% 
  group_by(ID) %>% 
  summarise(Percent_below_30c = sum(Temperature<30)/n()*100) %>% 
  filter(Percent_below_30c>80) %>% 
  select(ID) %>% 
  unlist() %>%
  str_sub(1,3) %>% 
  unique()

keep = keep[keep!=faulty_temp_measurements]
temperature_measurements %<>% filter(!(str_sub(ID,1,3) %in% faulty_temp_measurements))

cat(glue("<ul><b>Participants:</b>
          <li>We have 106 temperature measurements</li>
          <li>53 for skin and 53 for air temperature</li>
          <li>In those 53 skin temperature files we have a total of {length(keepall)} participants</li>
          <li>We drop {length(nokeep)} participants because they dont have temperature measurements in both settings ({length(keepall)-length(nokeep)} participants left) 
          <li>We drop 1 participant because of invalid biomarker measurements ({length(keep)+length(faulty_temp_measurements)} participants left)</li>
          <li>We drop {length(faulty_temp_measurements)} participant because of faulty skin temperature measurements ({length(keep)} participants left)</li></ul>"))

rm(total,nokeep,keepall,faulty_temp_measurements)
```

```{r baseline-characteristics}
baseline_characteristics <- basic %>% full_join(baseline, by="Code") %>% full_join(chrono, by="Code")

baseline_characteristics %<>% select("Group","Code","Age","Height","Weight","Sex","education_level","Smoking_status",
                          "alcohol_freq","Physical_exercise","screen_hours_day",
                          "First_day_in_village", "Last_day_in_village","First_Sampling_day","Second_Sampling_day")

# change variable types into factors
baseline_characteristics %<>% mutate(across(c(Group,Sex,education_level,Smoking_status,alcohol_freq, 
                                   Physical_exercise),~as_factor(.)))

# change variable types into numeric
baseline_characteristics %<>% mutate(across(Age,~as.numeric(.)))

# format dates
baseline_characteristics %<>% 
  mutate(across(c("First_day_in_village", "Last_day_in_village", 
                  "First_Sampling_day","Second_Sampling_day"),
                ~as_date(., format="%d/%m/%Y")))

# calculate BMI
baseline_characteristics %<>% mutate(BMI = bmi(Weight,(Height/100)))

# create BMI categories: 
# underweight under 18.5 kg/m2, normal weight: 18.5 to 25, overweight: 25 to 30
baseline_characteristics %<>% 
  mutate(BMIcat = cut(BMI,
                      breaks=c(-Inf, 18.5, 25, +Inf),
                      labels=c("underweight","normal weight","overweight")))

# revalue categorical variables to their actual values: 
baseline_characteristics$Group %<>% recode_factor("1"="First urban", "2"="First mountainous")

baseline_characteristics$education_level %<>% recode_factor( "2"="Secondary", 
                                                  "3"="University/college", 
                                                  "4"="Master/PhD")

baseline_characteristics$Smoking_status %<>% recode_factor("1"="Smoker", 
                                                "2"="Non-smoker", 
                                                "3"="Former smoker")

baseline_characteristics$alcohol_freq %<>% recode_factor("2"="Weekly",
                                            "3"="Monthly",
                                            "4"="Rarely/Never")

baseline_characteristics$Physical_exercise %<>% recode_factor("1"="Yes", "2"="No")

baseline_characteristics$Sex %<>% recode_factor("1"="Female", "2"="Male")

# calculate washout period based on group 
# for Group 2: use of 5 days in order to account the 5-day min stay in urban setting
baseline_characteristics %<>% mutate(washout_days = as.numeric(if_else(Group=="First mountainous",
                                                                       Second_Sampling_day - 5 - Last_day_in_village,
                                                                       First_day_in_village - First_Sampling_day)))

# calculate days in village based on group 
baseline_characteristics %<>% mutate(days_mountain = as.numeric(if_else(Group=="First mountainous",
                                                                        First_Sampling_day - First_day_in_village,
                                                                        Second_Sampling_day - First_day_in_village)))

baseline_characteristics %<>% select(!c(First_Sampling_day,First_day_in_village,Second_Sampling_day,Last_day_in_village))

# keep only the codes we using for the analysis
baseline_characteristics %<>% filter(Code %in% (temperature_measurements$ID %>% str_sub(1,3)))

pvalue <- function(x, ...) {
    # Construct vectors of data y, and groups (strata) g
    x <- x[c(1,2)] # drop ovreall
    y <- unlist(x)
    g <- factor(rep(1:length(x), times=sapply(x, length)))
    if (is.numeric(y)) {
        # For numeric variables, perform a standard 2-sample t-test
        p <- t.test(y ~ g)$p.value
    } else {
        # For categorical variables, perform a fisher test of independence
        p <- fisher.test(table(y, g))$p.value
    }
    # Format the p-value, using an HTML entity for the less-than sign.
    # The initial empty string places the output on the line below the variable label.
    c("", sub("<", "&lt;", format.pval(p, digits=3, eps=0.001)))
}

table1(~ Age + Sex + BMI + BMIcat + education_level + Smoking_status + alcohol_freq +
         Physical_exercise + screen_hours_day + days_mountain + washout_days | Group,
         data = baseline_characteristics , extra.col=list(`P-value`=pvalue))

rm(basic,chrono,baseline,pvalue)
```

```{r temperature-data-pre-processing}
temperature_measurements %<>% 
  inner_join(code_setting, by=c("ID"="Code_order"))                             %>% # add setting column
  mutate(ID=glue("{str_sub(ID,1,4)}{if_else(setting==\"Urban\",\"U\",\"M\")}")) %>% # fix id so it contains setting information
  arrange(desc(setting))                                                        %>% # short by setting
  mutate(ID=as_factor(ID), setting=as_factor(setting))                          %>% # ID as factor
  mutate(Date=dmy_hm(Date))                                                     %>% # fix date format
  na.omit()

rm(code_setting)
```

```{r temperature-plots}
p1 <-temperature_measurements %>%
  filter(setting=="Urban") %>%
  ggplot(aes(x=hour(Date),y=Temperature)) + 
  facet_wrap(vars(ID)) + 
  geom_line() + 
  ggtitle("Temperature plots Urban") +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank())

p2 <-temperature_measurements %>% 
  filter(setting=="Mountainous") %>%
  ggplot(aes(x=hour(Date),y=Temperature)) + 
  facet_wrap(vars(ID)) + 
  geom_line() + 
  ggtitle("Temperature plots Mountainous") +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank())

ggarrange(p1,p2)
rm(p1,p2)
```

```{r rqa}
rqa_rad <- 0.01

rqa_measurements <-
  temperature_measurements %>%
  group_by(ID,setting) %>% 
  summarise(rqa(time.series = Temperature, time.lag = 1, radius = rqa_rad, embedding.dim=2)[2:13] %>% as_tibble(), .groups = "keep") %>% 
  ungroup() %>%
  select(!c(Lmax,Vmax,LmeanWithoutMain,RATIO))

kable(rqa_measurements, caption = glue("RQA analysis Settings: time.lag=1, radius = {rqa_rad}"))
```
## RQA Plots
```{r generate-rqa-plots, include=FALSE, eval=TRUE}
i <- 1
plots <- list()

for (participant in keep) {
  myrqadataU <- temperature_measurements %>% filter(ID==glue("{participant}-U")) %>% mutate(observations = 1:n())
  myrqadataM <- temperature_measurements %>% filter(ID==glue("{participant}-M")) %>% mutate(observations = 1:n())
  
  maxU <- max(myrqadataU$observations)
  maxM <- max(myrqadataM$observations)
  
  p1 <- recurrencePlot(time.series = myrqadataU$Temperature, time.lag = 1, 
                       radius = rqa_rad, embedding.dim = 2, main=glue("{participant} Urban"),
                       scales = list(y = list(at = myrqadataU$observations[c(maxU-2,maxU %/% 2, 1)], 
                                              label = myrqadataU$Date[c(maxU-2,maxU %/% 2,1)]),
                                     x = list(at = myrqadataU$observations[c(1,maxU %/% 2,maxU-2)], 
                                              label = myrqadataU %$% glue("{hour(Date)}:{minute(Date)}")[c(1,maxU %/% 2,maxU-2)])),
                       ylab="",xlab="", plot=FALSE)

  plots[[i]] <- p1
  i <- i + 1
  
  p2 <- recurrencePlot(time.series = myrqadataM$Temperature, time.lag = 1, 
                       radius = rqa_rad, embedding.dim = 2, main=glue("{participant} Mountainous"),
                       scales = list(y = list(at = myrqadataM$observations[c(maxM-2,maxM %/% 2, 1)], 
                                              label = myrqadataM$Date[c(maxU-2,maxM %/% 2,1)]),
                                     x = list(at = myrqadataM$observations[c(1,maxM %/% 2,maxM-2)], 
                                              label = myrqadataM %$% glue("{hour(Date)}:{minute(Date)}")[c(1,maxM %/% 2,maxM-2)])),
                       ylab="",xlab="")

  plots[[i]] <- p2
  i <- i + 1
}
```

```{r plot-rqa-plots, eval=TRUE}
for (i in seq(1,length(plots),2)) {
  grid.arrange(plots[[i]],plots[[i+1]], ncol=2,widths = c(20,20))
}
```
<br>
```{r clean-up}
rm(plots,i,p1,p2,maxM,maxU,rqa_rad,participant,myrqadataM,myrqadataU,keep,temperature_measurements)
```
## PCA analysis
```{r pca-analysis}
res_pca <- PCA(rqa_measurements %>% select(!setting) %>% column_to_rownames("ID"), graph = FALSE)
kable(get_eig(res_pca), caption = "Eigenvalues/variances")

fviz_screeplot(res_pca, addlabels = TRUE, ylim = c(0, 50))

kable(res_pca$var$coord,   caption = "Coordinates of variables")
kable(res_pca$var$contrib, caption = "Contribution of variables")

fviz_pca_var(res_pca, col.var="contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)

# Contributions of variables to PC1
fviz_contrib(res_pca, choice = "var", axes = 1, top = 10)
# Contributions of variables to PC2
fviz_contrib(res_pca, choice = "var", axes = 2, top = 10)
# Extract the results for individuals

fviz_pca_ind(res_pca, col.ind = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE ,# Avoid text overlapping (slow if many points)
             title = "Quality of the individuals on the factor map")

fviz_pca_ind(res_pca,
             label = "none", # hide individual labels
             habillage = rqa_measurements$setting, # color by groups
             palette = c("#00AFBB", "#E7B800", "#FC4E07"),
             addEllipses = TRUE # Concentration ellipses
)
```
## RQA descriptives Overall and by setting
```{r rqa-descriptives}
# add dim1 dim2
model_data <- rqa_measurements %>% inner_join(res_pca$ind$coord %>% as_tibble(rownames = NA) %>% rownames_to_column("ID") %>% select(ID,Dim.1,Dim.2), by="ID")

model_data %>% bind_rows(model_data %>% mutate(setting="Overall")) %>%
  select(!ID) %>% pivot_longer(!setting , names_to = "metric", values_to = "value") %>% group_by(metric,setting) %>%
  summarise(mean = mean(value),
            sd   = sd(value)  ,
            q0   = quantile(value,0),
            q25  = quantile(value,.25),
            q50  = quantile(value,.50),
            q75  = quantile(value,.75),
            q90  = quantile(value,.90),
            q95  = quantile(value,.95),
            q100 = quantile(value,1), .groups = "keep") %>% kable()

rm(res_pca,rqa_measurements)
```
## Paired T test after log transformation for each variable for the 2 groups (alternative hypothesis: true difference in means is not equal to 0)
```{r paired-t-test}
kable(model_data %>%
  select_if(is.numeric) %>%
  select(!`Dim.2`) %>%
  mutate(across(REC:Vmean, ~log1p(.)),`Dim.1` = log1p(`Dim.1` + abs(min(`Dim.1`)))) %>%
  map_df(~ broom::tidy(t.test(. ~ model_data$setting, paired = TRUE) ), .id = 'metric'), caption = "Paired - T test")
```
## Leptin,cortisol and adiponectin are adjusted for creatinine and then log transformed (Metrics are log transformed in a previous step)
```{r model-data-log}
measurements$creatinine %<>% {if_else(.<0,0.1,.)}
measurements %<>% mutate(across(!c(ID,creatinine, SampleType),~log1p(./creatinine))) %>% select(!creatinine)
model_data %<>% mutate(across(REC:Vmean, ~log1p(.)),`Dim.1` = log1p(`Dim.1` + abs(min(`Dim.1`)))) %>% select(!`Dim.2`)

model_data %<>% 
  inner_join(measurements, by=c("ID")) %>% 
  mutate(across(!c(ID,setting,SampleType),~scale(.)[TRUE])) %>% 
  ungroup() %>%
  mutate(ID=str_sub(ID,1,3)) %>%
  right_join(baseline_characteristics %>% select(Code,Age,Sex),by=c("ID"="Code")) %>%
  pivot_longer(REC:`Dim.1`, names_to="MesurmentName",values_to="Mesurment") %>%
  nest_by(MesurmentName)
```
## Leptin models
```{r leptin-models-log}
model_formula <- as.formula(leptin ~ Mesurment + SampleType + Age + (1|ID))
print(glue("##### Formula:{as.character(model_formula)} <br> Number of observations:{nobs(lmer(model_formula, data=model_data$data[[1]]))}"))
model_leptin <- model_data %>% summarise(tidy(lmer(model_formula, data=data),conf.int=TRUE),.groups="keep") %>% ungroup()
model_leptin %<>% filter(term=="Mesurment")  %>% arrange(p.value)
kable(model_leptin %>% select(!group), caption = "p.values values for Metabolites (14 total models)")
```
## Adiponectin models
```{r adiponectin-models-log}
# adiponectin models
model_formula <- as.formula(adiponectin ~ Mesurment + SampleType + Age + (1|ID))
print(glue("##### Formula:{as.character(model_formula)} <br> Number of observations:{nobs(lmer(model_formula, data=model_data$data[[1]]))}"))
model_adiponectin <- model_data %>% summarise(tidy(lmer(model_formula, data=data),conf.int=TRUE),.groups="keep") %>% ungroup()
model_adiponectin %<>% filter(term=="Mesurment") %>% arrange(p.value)
kable(model_adiponectin %>% select(!group), caption = "p.values values for Metabolites (14 total models)")
```
## Cortisol models
```{r cortisol-models-log}
model_formula <- as.formula(cortisol ~ Mesurment + SampleType + Age + (1|ID))
print(glue("##### Formula:{as.character(model_formula)} <br> Number of observations:{nobs(lmer(model_formula, data=model_data$data[[1]]))}"))
model_cortisol <- model_data %>% summarise(tidy(lmer(model_formula, data=data),conf.int=TRUE),.groups="keep") %>% ungroup()
model_cortisol %<>% filter(term=="Mesurment") %>% arrange(p.value)
kable(model_cortisol %>% select(!group), caption = "p.values values for Metabolites (14 total models)")
```
## Metrics ~ setting models
```{r metrics-models}
model_formula <- as.formula(Mesurment ~ setting + Age + Sex)
print(glue("##### Formula:{as.character(model_formula)} <br> Number of observations:{nobs(lm(model_formula, data=model_data$data[[1]]))}"))
model_metrics <- model_data %>% summarise(tidy(glm(model_formula, data=data),conf.int=TRUE),.groups="keep") %>% ungroup()
model_metrics %<>% filter(str_detect(term,"setting")) %>% arrange(p.value)
kable((model_metrics))
```